---
title: "US Ancestry Census Map"
subtitle: "Interactive Visualization of Ancestry Distribution Across States"
author: "Census Data Explorer"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
editor: visual
---

## Overview

This document presents an interactive map visualization of ancestry distribution across the United States based on Census data. The map allows users to explore different ancestry groups and their geographic patterns.

## Data Setup

First, let's load the necessary libraries and data:

```{r}
#| message: false
#| warning: false

library(sf)
library(leaflet)
library(dplyr)
library(tidyr)
library(RColorBrewer)

# Load geographic data
counties <- readRDS("data/counties.rds")
states <- readRDS("data/states.rds")

# Convert to sf if needed
if (!inherits(states, "sf")) {
  states <- st_as_sf(states)
}

# Create sample ancestry data
set.seed(42)
ancestry_data <- states %>%
  st_drop_geometry() %>%
  mutate(
    german = runif(n(), 10, 25),
    irish = runif(n(), 5, 18),
    english = runif(n(), 8, 20),
    italian = runif(n(), 2, 12),
    polish = runif(n(), 1, 8),
    french = runif(n(), 2, 6),
    mexican = runif(n(), 2, 35),
    chinese = runif(n(), 0.5, 4),
    african_american = runif(n(), 2, 40),
    asian = runif(n(), 1, 10)
  ) %>%
  select(GEOID, german, irish, english, italian, polish, french, mexican, chinese, african_american, asian)

# Join with geography
states_data <- states %>%
  left_join(ancestry_data, by = "GEOID")

head(states_data)
```

## Interactive Map: German Ancestry

```{r}
#| fig-width: 12
#| fig-height: 8

# Create color palette
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = c(0, max(states_data$german, na.rm = TRUE)),
  na.color = "#cccccc"
)

# Create labels
labels <- sprintf(
  "<strong>%s</strong><br/>German: %.1f%%",
  states_data$NAME,
  states_data$german
) %>%
  lapply(HTML)

# Create map
leaflet(data = states_data) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = -95, lat = 37.5, zoom = 4) %>%
  addPolygons(
    fillColor = ~pal(german),
    weight = 0.5,
    opacity = 0.5,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~german,
    opacity = 0.7,
    title = "German %",
    position = "bottomright"
  )
```

## Interactive Map: Mexican Ancestry

```{r}
#| fig-width: 12
#| fig-height: 8

# Create color palette for Mexican ancestry
pal_mexican <- colorNumeric(
  palette = "Blues",
  domain = c(0, max(states_data$mexican, na.rm = TRUE)),
  na.color = "#cccccc"
)

# Create labels
labels_mexican <- sprintf(
  "<strong>%s</strong><br/>Mexican: %.1f%%",
  states_data$NAME,
  states_data$mexican
) %>%
  lapply(HTML)

# Create map
leaflet(data = states_data) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = -95, lat = 37.5, zoom = 4) %>%
  addPolygons(
    fillColor = ~pal_mexican(mexican),
    weight = 0.5,
    opacity = 0.5,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = labels_mexican,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal_mexican,
    values = ~mexican,
    opacity = 0.7,
    title = "Mexican %",
    position = "bottomright"
  )
```

## Summary Statistics

Here are summary statistics for all ancestry groups:

```{r}
ancestry_summary <- states_data %>%
  st_drop_geometry() %>%
  select(german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  summarise(across(everything(), list(
    mean = mean,
    min = min,
    max = max,
    sd = sd
  ))) %>%
  round(2)

# Format for display
summary_long <- ancestry_summary %>%
  pivot_longer(everything()) %>%
  separate(name, into = c("ancestry", "stat"), sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value)

summary_long
```

## Top States by Ancestry

### German Ancestry
```{r}
states_data %>%
  st_drop_geometry() %>%
  select(NAME, german) %>%
  arrange(desc(german)) %>%
  head(10) %>%
  knitr::kable()
```

### African American Ancestry
```{r}
states_data %>%
  st_drop_geometry() %>%
  select(NAME, african_american) %>%
  arrange(desc(african_american)) %>%
  head(10) %>%
  knitr::kable()
```

### Mexican Ancestry
```{r}
states_data %>%
  st_drop_geometry() %>%
  select(NAME, mexican) %>%
  arrange(desc(mexican)) %>%
  head(10) %>%
  knitr::kable()
```

## How to Use

1. **Interactive Maps**: Hover over states to see detailed information
2. **Zoom & Pan**: Use your mouse to explore specific regions
3. **Color Scale**: Darker colors indicate higher percentages
4. **Full Application**: Visit the Shiny app at `http://localhost:3838` for the complete interactive experience

## Notes

- **Sample Data**: This document uses simulated ancestry data for demonstration purposes
- **Real Data**: To use actual Census data, obtain an API key from [https://api.census.gov/data/key_signup.html](https://api.census.gov/data/key_signup.html)
- **Data Source**: Based on American Community Survey (ACS) 5-year estimates
- **Geographies**: State-level visualization; county-level data available in the Shiny app
