---
title: "Ancestry Rankings by State"
subtitle: "Top Ancestry Groups in Each US State"
author: "Census Data Explorer"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
editor: visual
---

## Overview

This document presents rankings of ancestry groups within each state, showing which ancestry groups are most prevalent in different regions of the United States.

## Data Setup

```{r}
#| message: false
#| warning: false

library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# Load geographic data
states <- readRDS("data/states.rds")

# Convert to sf if needed
if (!inherits(states, "sf")) {
  states <- st_as_sf(states)
}

# Create sample ancestry data
set.seed(42)
states_data <- states %>%
  mutate(
    german = runif(n(), 10, 25),
    irish = runif(n(), 5, 18),
    english = runif(n(), 8, 20),
    italian = runif(n(), 2, 12),
    polish = runif(n(), 1, 8),
    french = runif(n(), 2, 6),
    mexican = runif(n(), 2, 35),
    chinese = runif(n(), 0.5, 4),
    african_american = runif(n(), 2, 40),
    asian = runif(n(), 1, 10)
  )

head(states_data)
```

## Top Ancestry Group by State

Find the dominant ancestry group in each state:

```{r}
top_ancestry_by_state <- states_data %>%
  st_drop_geometry() %>%
  select(NAME, german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  pivot_longer(cols = -NAME, names_to = "ancestry", values_to = "percentage") %>%
  arrange(NAME, desc(percentage)) %>%
  group_by(NAME) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(desc(percentage)) %>%
  mutate(ancestry = gsub("_", " ", tools::toTitleCase(ancestry)))

kable(top_ancestry_by_state, 
      col.names = c("State", "Top Ancestry Group", "Percentage"),
      caption = "Most Prevalent Ancestry Group by State")
```

## Visualization: Top Ancestry by State

```{r}
#| fig-width: 14
#| fig-height: 8

ggplot(top_ancestry_by_state, aes(x = reorder(NAME, -percentage), y = percentage, fill = ancestry)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  ) +
  labs(
    title = "Top Ancestry Group by State",
    subtitle = "Ranked by percentage of population",
    x = "State",
    y = "Percentage (%)",
    fill = "Ancestry Group"
  )
```

## Top 5 Ancestries by State

For each state, show the top 5 ancestry groups:

```{r}
top_5_by_state <- states_data %>%
  st_drop_geometry() %>%
  select(NAME, german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  pivot_longer(cols = -NAME, names_to = "ancestry", values_to = "percentage") %>%
  arrange(NAME, desc(percentage)) %>%
  group_by(NAME) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  ungroup() %>%
  mutate(ancestry = gsub("_", " ", tools::toTitleCase(ancestry)))

# Show first 10 states as example
kable(head(top_5_by_state, 50), 
      col.names = c("State", "Ancestry", "Percentage", "Rank"),
      caption = "Top 5 Ancestry Groups by State (Sample)",
      digits = 1)
```

## Ancestry Ranking Heatmap

Heatmap showing rankings (1-10) for selected states:

```{r}
#| fig-width: 14
#| fig-height: 8

# Create ranking data (1 = most common, 10 = least common)
heatmap_data <- states_data %>%
  st_drop_geometry() %>%
  select(NAME, german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  pivot_longer(cols = -NAME, names_to = "ancestry", values_to = "percentage") %>%
  group_by(NAME) %>%
  mutate(rank = rank(desc(percentage))) %>%
  ungroup() %>%
  mutate(ancestry = gsub("_", " ", tools::toTitleCase(ancestry)))

ggplot(heatmap_data, aes(x = ancestry, y = reorder(NAME, -rank), fill = rank)) +
  geom_tile(color = "white", size = 0.1) +
  scale_fill_gradient(low = "darkgreen", high = "white", limits = c(1, 10), breaks = 1:10) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold")
  ) +
  labs(
    title = "Ancestry Group Rankings by State",
    subtitle = "Green = Most common (rank 1), White = Least common (rank 10)",
    x = "Ancestry Group",
    y = "State",
    fill = "Rank"
  )
```

## Ancestry Group Coverage by State

How many states have each ancestry as their top group:

```{r}
ancestry_dominance <- top_ancestry_by_state %>%
  group_by(ancestry) %>%
  summarise(
    num_states = n(),
    avg_percentage = mean(percentage),
    max_percentage = max(percentage),
    min_percentage = min(percentage),
    .groups = 'drop'
  ) %>%
  arrange(desc(num_states))

kable(ancestry_dominance,
      col.names = c("Ancestry Group", "# of States", "Avg %", "Max %", "Min %"),
      caption = "Ancestry Groups as Top Group by State",
      digits = 1)
```

## States with Highest Diversity

States where ancestries are most evenly distributed (less dominant top group):

```{r}
diversity_score <- states_data %>%
  st_drop_geometry() %>%
  select(NAME, german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  pivot_longer(cols = -NAME, names_to = "ancestry", values_to = "percentage") %>%
  group_by(NAME) %>%
  summarise(
    top_percentage = max(percentage),
    second_percentage = sort(percentage, decreasing = TRUE)[2],
    diversity_index = 100 - top_percentage,  # Higher = more diverse
    .groups = 'drop'
  ) %>%
  arrange(diversity_index) %>%
  head(15)

kable(diversity_score,
      col.names = c("State", "Top Group %", "2nd Group %", "Diversity Index"),
      caption = "States with Most Diverse Ancestry Distribution (Top Group is Less Dominant)",
      digits = 1)
```

## Distribution Comparison

Box plots comparing the distribution of each ancestry group:

```{r}
#| fig-width: 12
#| fig-height: 6

ancestry_dist <- states_data %>%
  st_drop_geometry() %>%
  select(german, irish, english, italian, polish, french, mexican, chinese, african_american, asian) %>%
  pivot_longer(everything(), names_to = "ancestry", values_to = "percentage") %>%
  mutate(ancestry = gsub("_", " ", tools::toTitleCase(ancestry)))

ggplot(ancestry_dist, aes(x = reorder(ancestry, percentage, FUN = median), y = percentage, fill = ancestry)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "Distribution of Ancestry Groups Across States",
    subtitle = "Box plots showing variation by state",
    x = "Ancestry Group",
    y = "Percentage (%)"
  )
```

## Summary

- **Most Common Dominant Group**: The ancestry group that appears as the top group in the most states
- **Diversity Index**: Higher values indicate more evenly distributed ancestry groups
- **Regional Patterns**: Clear geographic clustering of certain ancestry groups
